---
title: "Evaluating differential co-expression methods using dcanr"
author:
  - name: Dharmesh D Bhuva
    affiliation:
      - Bioinformatics Division, Walter and Eliza Hall Institute of Medical Research, Parkville, VIC 3052, Australia
      - School of Mathematics and Statistics, University of Melbourne, Parkville, VIC 3010, Australia
date: "`r Sys.Date()`"
package: dcanr
output:
  BiocStyle::html_document:
    toc_float: true
    fig_caption: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{2. DC method evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction
Along with implementations for several differential co-expression analysis
methods, this package provides an evaluation framework to benchmark methods.
Real data are rarely available for the evaluation of differential co-expression
methods as such experiments are difficult to perform and in almost all cases,
the true/gold-standard network is unknown. As such, simulations are the only
means to perform a rigorous comparison and evaluation. Data from 812 simulations
were generated with settings and parameters are described in **Bhuva et al.
(manuscript in preparation)** and results included here. Along with the data, a
suite of visualisation and performance metrics are implemented.

This vignette describes functions in this package that enable comparative
evaluation of inference methods. All methods implemented in the package can be
evaluated and novel methods can be integrated into the evaluation framework with
relative ease.

# Simulation setup used to create the data
A dynamical systems simulator was used to generate synthetic expression data
with differential associations. Regulatory networks of 150 genes were sampled
from a *S. cerevisiae* regulatory network. Perturbation experiments were
performed when simulating data to induce differential associations. Simulation
parameters and networks were sampled to produce 812 distinct simulations. The
dataset included in dcanr, `sim102`, is an example of one such simulation. Two
knock-downs were performed simultaneously but independent of each other,
therefore, some samples may have both knock-downs while other may have either
one or none. Details on the simulation procedure can be found in the **Bhuva et
al. (manuscript in preparation)**. Up to 500 observations are sampled in each
simulation.

# Download the full simulated dataset
As the simulation is computationally intensive, data from the 812 simulations
have been precomputed and are available at **LINK**. The downloaded file
contains a list of simulation results which includes `sim102`, packaged with
`dcanr`. Each simulation can be accessed as shown below.

```{r eval=FALSE}
simdata <- load('simdata_directory/simdata.rda')
sim10 <- simdata[[10]]
```

# Running a pipeline on a simulation
Evaluations in the package are performed by creating an analysis pipeline and
packaging this into a function. Three possible ways exist to perform this:

1. Using standard in-built pipelines
2. Using custom pipelines
3. Retrieving pre-computed results from the standard pipelines

All of the above are made possible using a single function `dcPipeline` and are processed
depending on the arguments specified.

## Standard pipelines
A standard pipeline runs the in-built inference methods with their default
parameters. All 4 steps of an analysis are performed in sequence as described in
their respective publications. To run a standard pipeline on a simulation,
simply pass in a simulation and a method name from `dcMethods()`.

```{r message=FALSE, warning=FALSE, fig.wide=TRUE, fig.asp=1, fig.cap='Differential network analysis on simulations. Top-left to bottom-right: The original regulatory network used for simulations; the true induced differential network that results from both ADR1 and UME6 knock-downs (KDs); the predicted ADR1 knock-down differential network, and; the predicted UME6 knock-down differential network. Reg and green edges in the regulatory network represent activation and repression respectively. Edges in the differential network are coloured based on the knock-down that results in their differential co-expression. True positives in the predicted network are coloured based on their knock-down while false positives are coloured grey'}
library(dcanr)

#load the data: a simulation
data(sim102) 
#run a standard pipeline with the z-score method
dcnets <- dcPipeline(sim102, dc.func = 'zscore')
#plot the source network, true differential network and inferred networks
op <- par(no.readonly = TRUE)
par(mfrow = c(2, 2))
plotSimNetwork(sim102, main = 'Regulatory network')
plotSimNetwork(sim102, what = 'association', main = 'True differential association network')
plot(dcnets$ADR1, main = 'ADR1 KD predicted network')
plot(dcnets$UME6, main = 'UME6 KD predicted network')
par(op)
```

Results from a pipeline are provided as a list of `igraph` objects representing
the inferred networks, one for each knock-down. True positive predictions are
coloured based on the colour of the knocked-down node and false positives are
coloured grey. The `plotSimNetwork` function can be used to retrieve and plot
the true differential networks and the source regulatory network.

Any additional parameters for intermediate steps in a standard pipeline can be
passed as shown in the example below. However, doing so will produce results
different from the precomputed results discussed in (Section \@ref(precomp)).

```{r}
#run a standard pipeline with the z-score method with custom params
dcnets_sp <- dcPipeline(sim102,
                        dc.func = 'zscore',
                        cor.method = 'spearman', #use Spearman's correlation
                        thresh = 0.2) #cut-off for creating the network
```

## Custom pipelines
The most common use case of an evaluation framework is to benchmark new methods.
As such, this framework allows seamless integration of new methods. This is made
possible in `dcanr` by providing a single function to the `dcPipeline` method.
This function should follow the skeleton below:

```{r eval=FALSE}
#emat, a named matrix with samples along the columns and genes along the rows
#condition, a binary named vector consisiting of 1's and 2's
#returns a named adjacency matrix or an igraph object
myInference <- function(emat, condition, ...) {
  #your code here
  return(dcnet)
}
```

The following code shows the function used to call an in-built pipeline and how
a custom pipeline can be run.

```{r}
#custom pipeline function
analysisInbuilt <- function(emat, condition, dc.method = 'zscore', ...) {
  #compute scores
  score = dcScore(emat, condition, dc.method, ...)
  #perform statistical test
  pvals = dcTest(score, emat, condition, ...)
  #adjust tests for multiple testing
  adjp = dcAdjust(pvals, ...)
  #threshold and generate network
  dcnet = dcNetwork(score, adjp, ...)

  return(dcnet)
}

#call the custom pipeline
custom_nets <- dcPipeline(sim102, dc.func = analysisInbuilt)
```

## Retrieving pre-computed results {#precomp}
Following evaluation of a novel method, it is generally of interest to compare
how it performs against previous methods. Re-computing predictions for the 10
methods implemented in this package can be time consuming. As such, all
simulations come packaged with results from applying a standard pipeline using
each inference method. Pre-computed results are also retrieved using the
`dcPipeline` function, with the `precomputed` argument set to `TRUE`.

```{r message=FALSE, warning=FALSE}
#retrieve results of applying all available methods
allnets <- lapply(dcMethods(), function(m) {
  dcPipeline(sim102, dc.func = m, precomputed = TRUE)
})
names(allnets) <- dcMethods() #name the results based on methods

#get the size of the UME6 KD differential network
netsizes <- lapply(allnets, function(net) {
  length(igraph::E(net$UME6))
})
print(unlist(netsizes))
```

# Evaluate a pipeline
After inferring the differential co-expression network, the final step is to
evaluate performance based on the truth. True differential networks are computed
by performing a sensitivity analysis as described in the **Bhuva et al.
(manuscript in preparation)**. Three levels of true differential networks are
generated:

1. Direct - this represents direct TF-target regulatory interactions that are
affected by the knock-down (condition)
2. Influence - this network includes upstream TF-target interactions which are
indirect but causative
3. Association - this network includes all associations in the influence network
but adds all non-causative associations that are differential when conditioned
on the knock-down

All three levels of truth are included within the simulated data. Predicted
networks can be evaluated against any of these. We recommend using the
differential association network.

Along with the true differential co-expression network, a metric is required to summarise
performance of a method. The package provides 6 different metrics which can be accessed
by calling `perfMethods()`.

```{r}
#available performance metrics
print(perfMethods())
```

Evaluation is performed using the `dcEvaluate` function. We will run this with the 
precomputed inferences as described in (Section \@ref(precomp)). The F1-measure will
be computed to assess performance.

```{r message=FALSE, warning=FALSE}
#compute the F1-measure for the prediction made by each method
f1_scores <- lapply(allnets, function (net) {
  dcEvaluate(sim102, net, truth.type = 'association', combine = TRUE)
})
print(sort(unlist(f1_scores), decreasing = TRUE))
#compute the Matthew's correlation coefficient of the z-score inference
z_mcc <- dcEvaluate(sim102, dcnets, perf.method = 'MCC')
print(z_mcc)
```

These metrics can be combined to learn more about the advantages and limitations
of different methods. For instance, the precision and recall characteristics of
the different methods can be computed. For sim102, this shows that the
entropy-based method excels at recall while the z-score based method has high
precision.

```{r message=FALSE, warning=FALSE}
#compute precision
dcprec <- lapply(allnets, function (net) {
  dcEvaluate(sim102, net, perf.method = 'precision')
})
#compute recall
dcrecall <- lapply(allnets, function (net) {
  dcEvaluate(sim102, net, perf.method = 'recall')
})
```

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.small=TRUE}
#plot the precision and recall
plot(
  unlist(dcprec),
  unlist(dcrecall),
  type = 'n',
  main = 'Precision v recall',
  xlim = c(-0.25, 1),
  ylim = c(0, 1)
  )
text(
  unlist(dcprec),
  unlist(dcrecall),
  labels = names(allnets),
  type = 'n',
  main = 'Precision vs recall',
  cex = 1.2
  )
```

# Session info {.unnumbered}
```{r sessionInfo, echo=FALSE}
sessionInfo()
```
