---
title: "DC method evaluation using dcanr"
author:
  - name: Dharmesh D Bhuva
    affiliation:
      - Bioinformatics Division, Walter and Eliza Hall Institute of Medical Research, Parkville, VIC 3052, Australia
      - School of Mathematics and Statistics, University of Melbourne, Parkville, VIC 3010, Australia
date: "`r Sys.Date()`"
package: dcanr
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{2. DC method evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction
Along with methods, the package provides an evaluation framework to benchmark 
methods. Real data is rarely available for the evaluation of differential co-expression 
methods as such experiments are difficult to perform. As such, simulations are the
only means to perform evaluation successfully. Data from 812 simulations was 
generated and the settings and parameters are described in **PAPER**. This
data has been simulated previously and the results are made available here.
Along with the data, a suite of visualisation and performance metrics are implemented.

This vignette describes the functions in this package that enable comparative
evaluation of inference methods. All methods within the package can be evaluated
with custom evaluations made possible where required.

# Simulation setup used to create the data
A dynamical systems simulator was used to simulate expression data with differential
associations. Regulatory networks of 150 genes were sampled from a S. Cerevisiae
regulatory network. Perturbation experiments were performed when simulating data
to induce differential associations. Simulation parameters and networks were
sampled therefore producing 812 distinct simulations. Data included in the package
`sim102` is an example of one such simulation where two knock-downs were performed
simultaneously but independently therefore some samples may have both knock-downs
whereas other may have either one or none. Details on the simulation procedure can
be found in the **PAPER**. Upto 500 observations are sampled in each simulation.

# Download the full simulated dataset
As the simulation is computationally intensive, data from the 812 simulations has
been precomputed and is available at **LINK**. The downloaded file contains a list
of simulation results such as `sim102` which is packaged with `dcanr`. Each
simulation can be accessed as shown below.

```{r eval=FALSE}
simdata <- load('simdata_directory/simdata.rda')
sim10 <- simdata[[10]]
```


# Running a pipeline on a simulation
## Standard pipelines
## Custom pipelines
## Retrieving precomputed results

# Evaluate a pipeline

# Session info {.unnumbered}
```{r sessionInfo, echo=FALSE}
sessionInfo()
```
